---
## Introduction to R

A great resource  is R for Data Science that is freely available online [here](https://r4ds.had.co.nz/) - I heavily rely on this in the lecture. And another good option YaRrr! the Pirates Guide to R is [here](https://bookdown.org/ndphillips/YaRrr/)

### CRAN & Mirrors 
R codes are stored in a series of servers across the world through Comprehensive R Archive Network (CRAN) - all of R related information is stored separately in each server (with the same exact information)

So, likely, the first thing you'll need to do is to use a mirror near you - a list of mirrors can be found here

Once you set it, it's done. No need to go back to this step unless for some reason you want to pull from a different server (maybe if you move, one that is closer to you)

```{r}
options("repos" = c(CRAN = "https://mirror.las.iastate.edu/CRAN/"))
```

We call packages through "library" Such as:

"library(package_name)"

You have to call the package that you want in your notebook.
In a moment, we're going to start working with tidyverse and dplyr.
But, just to see the importing of a package, let's try it here.

```{r}
library(plm)
```

If you want o install a package you can do so, like this:

`install.packages(dplyr)'

OR you can install through the R Studio browser - I'll show you that now.  

### Basic Commands

Before we get to working with data, let's do a quick over view of some basic commands

To change the working directory we use:

setwd(/your/working/directory/)

to find out where you are in your working directory we use:

`getwd()'

```{r}
getwd()
```

```{r}
setwd('/Users/mkaltenberg/Dropbox (Personal)/Pace/ECO585/R notebooks')
```

### Assignments

In R documentation they almost always use `<-'but `=' also works.

```{r}
x <- 'hi there!'
x
y= 'bye'

```

You can change a variable name just as easily be reassigning the variable name.

```{r}
(y <- seq(0, 10, 2))
```

### Import/Export Data

To read a csv file you use

read.csv(path_to_csv_file)

To save a csv file you use write.csv(df, path_to_csv_file)

```{r}
setwd('/Users/mkaltenberg/Dropbox (Personal)/Pace/ECO585/R notebooks/')
birthweight <- read.csv('birthweight_data_trial.csv')
```

Ok, now you. Go onto classes and download the file and import into R Studio now. If you have an error or other issues, share your screen and I can help you.

Exporting data is also easy: 
```{r}
write.csv(birthweight, 'birthweight_2.csv')
```

### Help function

you can always ask for documentation, but that function is: help()

```{r}
help('dplyr')
help('read.csv')
```
For packages, there is also a summary about the package and what it does with vignette

vignette('dplyr')

### Removing objects

Objects can clog up our RAM, especially if they are large datafiles. If you want to remove an object the function is: rm()

```{r}
rm(x,y)
```

You can also removing EVERYTHING in your environment with: rm(list = ls())

Just remember that it will delete everything and you'd have to import all of your data again


